---
title: "Best Practices"
subtitle: "Using the Diesel Exhaust in Miners Study (DEMS) Data"
author: "Nancy Carmona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        df_paged: true
---

# Setup
```{r setup, include=TRUE}
#-----setup----

#set knitr options
knitr::opts_chunk$set(
    echo = TRUE,
    cache = FALSE,
    cache.comments = FALSE,
    message = FALSE,
    warning = FALSE
)

# Clear workspace of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
   
}

# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed.
pacman::p_load(dplyr, ggplot2, knitr, Hmisc, tidyr, purrr, broom)


```

```{r set.file.paths}
#-----set file paths----

# create a variable for the project path
project_path <- getwd()

# create a variable for the data path
data_path <- file.path(project_path, "data")

# create directory, if one does not already exist
dir.create(data_path, showWarnings = FALSE, recursive = TRUE)

```

```{r read.data, eval=TRUE}
#-----read data----

# # Nancy? option to simplify? 
# DEMS <- readRDS(file.path(data_path, "DEMSCombinedPersonal.rds"))

# first, specify file name and path
DEMS_file <- "DEMSCombinedPersonal.rds"
DEMS_path <- file.path(data_path, DEMS_file)

# Only download the file if it's not already present
if (!file.exists(DEMS_path)) {
    url <- paste("https://staff.washington.edu/high/envh556/Datasets", 
                 DEMS_file, sep = '/')
    download.file(url = url, destfile = DEMS_path)
}

# Output a warning message if the file cannot be found
if (file.exists(DEMS_path)) {
    DEMS <- readRDS(file = DEMS_path)
} else warning(paste("Can't find", DEMS_file, "!"))

# remove temporary variables
rm(url, DEMS_file, DEMS_path)

```


# Goal

The goal of this `.Rmd` is to demonstrate the implementation of coding best practices. For our example, we'll use an example dataset from the "Diesel Exhaust in Miners Study (DEMS)" borrowed from ENVH 556.

---

# Data Description

The underlying data to be used in the lab is from the DEMS study and is
described in [Coble et al 2010](https://academic.oup.com/annweh/article/54/7/747/202635).  We are focusing on the personal data collected
by NIOSH between 1998 and 2001.  In this lab we will mostly use the `ecdata`
variable which is a measurement of exposure to respirable elemental carbon, also
called REC.


## Data Exploration

Check the data out at a high level.

```{r data.description.1}
#-----data description 1---------- 

# check class
class(DEMS) 

# get dataframe dimensions
dim(DEMS) 

# dataframe varible names
names(DEMS) 

# This gives us details on the type of each variable in the dataset
sapply(DEMS, class) 

```

# Data Wrangling 

```{r data.description.2}
#-----data description 2----

# make DEMS a tibble
DEMS <- as_tibble(DEMS) %>%

  # focus on the key exposure data
  select(facilityno, u_s, ecdata, nodata, no2data) %>% 
  
  # data transformations
  mutate(ln_ecdata = log(ecdata), 
         log10_ecdata = log10(ecdata), 
         u_s = case_when(u_s == "u" ~ "underground", 
                         u_s == "s" ~ "surface", 
                         u_s == "m" ~ "mixed"), 
         u_s = factor(u_s, levels = c("surface", "mixed", "underground"))
         )

# first few rows and ALL variables (columns):
head(DEMS) 

# a basic summary of each variable
summary(DEMS) 

# a different basic summary of each variable from `Hmisc`
describe(DEMS) 

# tallies of the number of observations by facilityno
xtabs( ~ facilityno, data = DEMS) 

```

# Basic Data Description

Using the `tidyverse` `dplyr` package.

```{r table.with.dplyr}
#-----table with dplyr------

# create summary dataframe 
facility_ec <- DEMS %>% 
  group_by(facilityno) %>% 
  summarise(N = sum(!is.na(ecdata)),
            Nmiss = sum(is.na(ecdata)),
            mean = mean(ecdata, na.rm = TRUE),
            sd = sd(ecdata, na.rm = TRUE),
            se = sd/sqrt(N), 
            .groups = "drop" ) %>% 
  mutate_if(is.double, round, digits = 2)

# show tibble
facility_ec

# And here is the same result printed using kable
kable(facility_ec)
```

# Data Inference 

```{r regression.modeling.tidy}
no2_mods <- DEMS %>% 
  
  # group by facility
  group_by(facilityid) %>%
  
  # sum missing & filter out facilities with all missing data
  mutate(N = sum(!is.na(no2data))) %>% 
  filter(N != 0) %>% 
  
  # nest the tibble
  nest() %>% 
  
  # run regression model for each group, clean up with `tidy()`
  mutate(model = map(data, ~lm(ecdata ~ no2data, data = .x)), 
         tidied = map(model, tidy) 
         ) %>% 
  
  # unnest
  unnest(tidied) %>% 
  
  # filter to just NO2, our term of interest
  filter(term == "no2data")

```


## Data Visualization 

```{r hist.in.tidyverse, warning=FALSE, message=FALSE}
#-----hist in tidyverse---------

# histogram with density base plot
p <- ggplot(data=DEMS, aes(ecdata, na.rm=TRUE)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="white", binwidth=20) + 
  geom_density(alpha=.2, fill="red") + 
  labs(x = "REC (ug/m3)") +
  scale_y_continuous(expand = expand_scale(mult = c(0, .1))) +
  theme_classic()

p
```

```{r scatterplot.in.tidyverse, echo=TRUE, warning=FALSE, message=FALSE}
#-----scatterplot in tidyverse------------

# Plot facets with free scales. 
# Note 1:  the smoother for facility E is behaving strangely because so many 
# of the values are equal to zero. 
# Note 2: with facets, you see facility D21 has no `nodata`
ggplot(data = DEMS, aes(nodata, ecdata, color = u_s)) +
    geom_point() +
    facet_wrap(~ facilityno, nrow = 2, scales = "free") + 
    geom_smooth(se = FALSE) +
    labs(
    title = "Scatterplot of the DEMS NO (ppm) vs. REC (ug/m3) data \nby facility, colored by measurement location",
    x = "NO (ppm)",
    y = "REC (ug/m3)",
    color = "Measurement\nlocation"
    ) + 
  scale_color_brewer(palette = "Dark2") +
  theme_bw()
    
```

# Session Information, Code, and Functions

The next three chunks should be included in the appendix of every R Markdown so 
that you document your session information, code, and functions defined in the 
document. This supports the reproducibility of your work.

```{r session.info}
#-----session information: beginning of Appendix -----------

# This promotes reproducibility by documenting the version of R and every package
# you used.
sessionInfo()

```

```{r appendix, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE, include=TRUE}
#-----appendix------------
```

```{r functions.defined.in.this.Rmd, eval = TRUE}
#-----functions defined in this Rmd ------------

# Show the names of all functions defined in the .Rmd
# (e.g. loaded in the environment)
lsf.str()

# Show the definitions of all functions loaded into the current environment  
lapply(c(lsf.str()), getAnywhere)

```
