---
title: "Worst Practices"
subtitle: "Using the Diesel Exhaust in Miners Study (DEMS) Data"
author: "Nancy Carmona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        df_paged: true
---

# Setup
```{r setup, include=TRUE}
#-----setup----

#set knitr options
knitr::opts_chunk$set(
    echo = TRUE,
    cache = FALSE,
    cache.comments = FALSE,
    message = FALSE,
    warning = FALSE
)

# Clear workspace of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
   
}

# Load packages
library(ggplot2)
library(Hmisc)
library(dplyr)
library(knitr)
library(lubridate)

### Some of the worst practices this chunk ### 
# Load all packages at once 
# Assuming all packages were previously installed 
# Loading packages that are not needed 
# Not loading all packages that will be used in the analysis 

```

```{r read.data, eval=TRUE}
#-----read data----

# read data 
  DEMS <- DEMSCombinedPersonal <- readRDS("~/Coders group/Prep/nancy_best_practice/data/DEMSCombinedPersonal.rds")

# first, specify file name and path
DEMS_file <- "DEMSCombinedPersonal.rds"
DEMS_path <- file.path(data_path, DEMS_file)

# Only download the file if it's not already present
if (!file.exists(DEMS_path)) {
    url <- paste("https://staff.washington.edu/high/envh556/Datasets", 
                 DEMS_file, sep = '/')
    download.file(url = url, destfile = DEMS_path)
}

# Output a warning message if the file cannot be found
if (file.exists(DEMS_path)) {
    DEMS <- readRDS(file = DEMS_path)
} else warning(paste("Can't find", DEMS_file, "!"))

# remove temporary variables
#rm(url, DEMS_file, DEMS_path)


### Some of the worst practices this chunk ### 
# not using file.path() to construct directories 
# read data from absolute path instead of using directories 
# plat-form specific paths (Mac vs. Windows)
# large objects should be removed to reduce memory being used 

```


# Goal

The goal of this `.Rmd` is to demonstrate the implementation of coding best practices. For our example, we'll use an example dataset from the "Diesel Exhaust in Miners Study (DEMS)" borrowed from ENVH 556.

---

# Data Description

The underlying data to be used in the lab is from the DEMS study and is
described in [Coble et al 2010](https://academic.oup.com/annweh/article/54/7/747/202635).  We are focusing on the personal data collected
by NIOSH between 1998 and 2001.  In this lab we will mostly use the `ecdata`
variable which is a measurement of exposure to respirable elemental carbon, also
called REC.


## Data Exploration

Check the data out at a high level.

```{r data.description.1}
#-----data description 1---------- 

class(DEMS)  # check class

dim(DEMS) # get dataframe dimensions

names(DEMS) # dataframe varible names


sapply(DEMS, class) # This gives us details on the type of each variable in the dataset

View(DEMS) 
# view data

 
### Some of the worst practices this chunk ### 
# commenting at end of lines or after the code it refers to 
# having a large chunk instead separating data descriptions 
# combined "Data Wranging" and "Data Exploration" 
# adding comments at the end of lines or after the code 
# including "View" in the script 

#-----data description 2----

# make DEMS a tibble
DEMS <- as_tibble(DEMS) %>%

  # focus on the key exposure data
  select(facilityno, u_s, ecdata, nodata, no2data) %>% 
  
  # data transformations
  mutate(ln_ecdata = log(ecdata), 
         log10_ecdata = log10(ecdata), 
         u_s = case_when(u_s == "u" ~ "underground", 
                         u_s == "s" ~ "surface", 
                         u_s == "m" ~ "mixed"), 
         u_s = factor(u_s, levels = c("surface", "mixed", "underground"))
         )

# first few rows and ALL variables (columns):
head(DEMS) 

# a basic summary of each variable
summary(DEMS) 

# a different basic summary of each variable from `Hmisc`
describe(DEMS) 

# tallies of the number of observations by facilityno
xtabs( ~ facilityno, data = DEMS) 


```

# Basic Data Description


```{r table.with.baseR}
# make dataframes for different facilities
DEMSa <- filter(DEMS, facilityid == "A")
DEMSb <- filter(DEMS, facilityid == "B")
DEMSc <- filter(DEMS, facilityid == "C")
DEMSd <- filter(DEMS, facilityid == "D")
DEMSg <- filter(DEMS, facilityid == "D")
DEMSh <- filter(DEMS, facilityid == "H")
DEMSi <- filter(DEMS, facilityid == "I")

# calculate stats for facility A
a_N <- dim(DEMSa)[1]
a_miss <- sum(is.na(DEMSa$ecdata))
a_mean <- mean(DEMSa$ecdata, na.rm = TRUE)
a_sd <- sd(DEMSa$ecdata, na.rm = TRUE)

# calculate stats for facility B
b_N <- dim(DEMSb)[1]
b_miss <- sum(is.na(DEMSb$ecdata))
b_mean <- mean(DEMSa$ecdata, na.rm = TRUE)
b_sd <- sd(DEMSb$ecdata, na.rm = TRUE)

# calculate stats for facility C
c_N <- dim(DEMSb)[1]
c_miss <- sum(is.na(DEMSb$ecdata))
c_mean <- mean(DEMSa$ecdata, na.rm = TRUE)
c_sd <- sd(DEMSb$ecdata, na.rm = TRUE)

# caluate for facility D
d_N <- dim(DEMSd)[1]
d_miss <- sum(is.na(DEMSd$ecdata))
d_mean <- mean(DEMSd$ecdata, na.rm = TRUE)
d_sd <- sd(DEMSd$ecdata, na.rm = TRUE)

# calculate stats for facility E
g_N <- dim(DEMSg)[1]
g_miss <- sum(is.na(DEMSg$ecdata))
g_mean <- mean(DEMSg$ecdata, na.rm = TRUE)
g_sd <- sd(DEMSg$ecdata, na.rm = TRUE)

# calculate stats for facility H
h_N <- dim(DEMSh)[1]
h_miss <- sum(is.na(DEMSg$ecdata))
h_mean <- mean(DEMSg$ecdata, na.rm = TRUE)
h_sd <- sd(DEMSh$ecdata, na.rm = TRUE)

# calculate stats for facility I
i_N <- dim(DEMSi)[1]
i_miss <- sum(is.na(DEMSi$ecdata))
i_mean <- mean(DEMSi$ecdata, na.rm = TRUE)
i_sd <- sd(DEMSi$ecdata, na.rm = TRUE)

# you've still got to put this all together?! 


### Some of the worst practices this chunk ### 
# easier to make mistakes when we copy and paste 
# it is easier to use tidyverse and dplyr packages 
# using group_by() and summarise() this code is much simpler 

```

 
# Data Inference 

```{r regression.modeling.baseR}
# linear regression for each model 
a_mod <- summary(lm(data = DEMSa, ecdata ~ no2data))$coefficients
b_mod <- summary(lm(data = DEMSb, ecdata ~ no2data))$coefficients
e_mod <- summary(lm(data = DEMSe, ecdata ~ no2data))$coefficients
g_mod <- summary(lm(data = DEMSg, ecdata ~ no2data))$coefficients
h_mod <- summary(lm(data = DEMSh, ecdata ~ no2data))$coefficients
i_mod <- summary(lm(data = DEMSi, ecdata ~ no2data))$coefficients

### Some of the worst practices this chunk ### 
# massive subsetting into new variables instead of using group_by
# using a lot of copy/paste and repetition
# not using functions or pipelines to reduce copy/paste
```

 

## Data Visualization 

```{r hist.in.tidyverse}

#-----hist in tidyverse---------
# load packages in random places 
library(ggplot2)

# histogram with density base plot
p <- ggplot(data=DEMS, aes(ecdata, na.rm=TRUE)) + geom_histogram(aes(y=..density..), colour="black", fill="white", binwidth=20) + geom_density(alpha=.2, fill="red") + labs(x = "REC (ug/m3)") +         scale_y_continuous(expand = expand_scale(mult = c(0, .1))) + theme_classic()

p

### Some of the worst practices this chunk ### 
# using poor code spacing, layout, indentation, etc., making code hard to read
# loading a package randomly 

```

```{r scatterplot.in.tidyverse, echo=TRUE }
 #-----scatterplot in tidyverse------------

# Plot facets with free scales. 
# Note 1:  the smoother for facility E is behaving strangely because so many 
# of the values are equal to zero. 
# Note 2: with facets, you see facility D21 has no `nodata`
ggplot(data = DEMS, aes(nodata, ecdata, color = u_s)) +
geom_point() +
facet_wrap(~ facilityno, nrow = 2, scales = "free") 
+  geom_smooth(se = FALSE) +  labs(
    title = "Scatterplot of the DEMS NO (ppm) vs. REC (ug/m3) data \n by facility, colored by measurement location",
    x = "NO (ppm)",
    y = "REC (ug/m3)",
    color = "Measurement\nlocation"
    ) + 
  scale_color_brewer(palette = "Dark2") +
  theme_bw()
    
### Some of the worst practices this chunk ### 
# not following + at beginning of each attribute 
# ignoring warnings 

```

# Session Information, Code, and Functions

The next three chunks should be included in the appendix of every R Markdown so 
that you document your session information, code, and functions defined in the 
document. This supports the reproducibility of your work.

```{r session.info}
#-----session information: beginning of Appendix -----------

# This promotes reproducibility by documenting the version of R and every package
# you used.
sessionInfo()

```

```{r appendix, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE, include=TRUE}
#-----appendix------------
```

```{r functions.defined.in.this.Rmd, eval = TRUE}
#-----functions defined in this Rmd ------------

# Show the names of all functions defined in the .Rmd
# (e.g. loaded in the environment)
lsf.str()

# Show the definitions of all functions loaded into the current environment  
lapply(c(lsf.str()), getAnywhere)

```
