---
title: "Ioslides Example"
author: "Brian High"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  ioslides_presentation:
    fig_caption: yes
    fig_retina: 1
    fig_width: 5
    fig_height: 3
    keep_md: true
    smaller: false
    incremental: false
    logo: img/logo_128.png
    css: inc/deohs-ioslides-theme.css
    template: inc/deohs-default-ioslides.html
---

```{r set_knitr_options, echo = FALSE}
suppressMessages(library(knitr))
opts_chunk$set(tidy = FALSE, cache = FALSE, echo = TRUE)
```

<!-- Note: Custom CSS changes title font and removes footer gradient. -->
<!-- Note: Custom HTML template replaces logo with banner on title page.-->

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Table

```{r iris, echo = TRUE}
library(kableExtra)
knitr::kable(head(iris)) %>% kable_styling(font_size = 18)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

## Slide with Crosstalk Elements {.smaller}

```{r, echo=FALSE, message=FALSE}
library(crosstalk)
library(leaflet)
library(DT)

# Wrap data frame in SharedData
sd <- SharedData$new(quakes[sample(nrow(quakes), 100),])

# Create a filter input
filter_slider("mag", "Magnitude", sd, column = ~ mag, step = 0.1, width = 400)

# Use SharedData like a dataframe with Crosstalk-enabled widgets
bscols(
  leaflet(sd, width = 400, height = 300) %>% addTiles() %>% addMarkers(),
  datatable(sd,
    extensions = "Scroller", style = "bootstrap", class = "compact",
    width = "100%", height = "30%",
    options = list(deferRender = TRUE, scrollY = 200, scroller = TRUE))
)
```

