---
title: "Shower Example"
author: "Brian High"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdshower::shower_presentation:
    self_contained: true
    katex: true
    ratio: 16x10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<style>
/* Disable page numbers */
.slide:after {
    display: none;
}
/* Increase font size of code and output */ 
code,pre {
    font-size: 20px;
}
</style>

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Table

```{r iris, echo = FALSE}
library(kableExtra)
knitr::kable(head(iris)) %>% kable_styling(font_size = 18)
```

## Slide with Plot

```{r pressure, echo =FALSE, fig.height = 4}
plot(pressure)
```

## Slide with Crosstalk Elements

<!-- Note: The DT::datatable element will be misformatted. -->

<div style="font-size: 65%">

```{r, echo=FALSE, message=FALSE}
library(crosstalk)
library(leaflet)
library(DT)

# Wrap data frame in SharedData
sd <- SharedData$new(quakes[sample(nrow(quakes), 100),])

# Create a filter input
filter_slider("mag", "Magnitude", sd, column = ~ mag, step = 0.1, width = 400)

# Use SharedData like a dataframe with Crosstalk-enabled widgets
bscols(
  leaflet(sd, width = 400, height = 275) %>% addTiles() %>% addMarkers(),
  datatable(sd,
    extensions = "Scroller", style = "bootstrap", class = "compact",
    width = "100%", height = "80%",
    options = list(deferRender = TRUE, scrollY = 150, scroller = TRUE))
)
```

</div>


