---
title: "Many Models with lme4::lmer() and broom.mixed::tidy()"
author: "Brian High"
date: "9/7/2020"
output:
  html_document:
    df_print: paged
    keep_md: yes
  pdf_document:
    fig_caption: yes
    keep_md: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives

- Run multiple models with `lmer()`, `tidy()` and `confint()` using `future_map()`.
- Compare `confint()` results with `tidy(conf.int = TRUE)` results.
- Use furrr package to provide `future_map()` for parallel processing.

## Setup

We will use the *broom.mixed* package to support `lmer()` from the *lme4* package.

```{r}
# Install pacman if not installed.
if (!require(pacman)) {
  install.packages('pacman', repos = 'http://cran.us.r-project.org')
}

# Load packages, installing as needed.
pacman::p_load(nycflights13,
               tibble,
               tidyr,
               dplyr,
               broom.mixed,
               modelr,
               purrr,
               furrr,
               lme4,
               tictoc)

# Show number of available CPU cores.
availableCores()

# Set number of multicore "workers" to 1/2 the number of cores.
plan(multiprocess, workers = availableCores()/2)
```

## Prepare formulas

We use a list so that we can have named groups of formulas.

```{r}
# Create a list of formulas for use with lmer().
formulas <- list(
  crude = c(
    "arr_delay ~ distance + (1|carrier)",
    "arr_delay ~ air_time + (1|carrier)",
    "arr_delay ~ month + (1|carrier)"
  ),
  min = c(
    "arr_delay ~ distance + air_time + (1|carrier)",
    "arr_delay ~ distance + month + (1|carrier)",
    "arr_delay ~ air_time + month + (1|carrier)"
  ),
  max = c("arr_delay ~ distance + air_time + month + (1|carrier)")
)

# Convert the list of formulas into a data.frame (tibble).
formula_df <- formulas %>%
  enframe(name = "fgroup", value = "formula") %>% unnest(formula)
```

## Prepare data

```{r}
# Select only the model variables to minimize parallelization overhead.
data_df <- 
  flights %>% select(arr_delay, distance, air_time, month, carrier)
```

## Run the models

We could use `nest()` for the following steps in a single pipeline, using 
`tidy()` for confidence intervals, but the terms might be misaligned. 

Instead, using a stepwise approach, we retain the term names, use `confint()` 
to get confidence intervals, join on the terms and the formula number, and 
show the results to compare the two methods of getting confidence intervals.

```{r, message=FALSE}
# Start timer.
tic()

# Fit the models with lmer() using future_map() for multicore processing.
model_fit_list <- formula_df$formula %>% future_map(lmer, data = data_df)

# Extract the estimates with broom.mixed::tidy().
est <- 
  model_fit_list %>%
  future_map(tidy, conf.int = TRUE, conf.level = 0.95) %>%
  bind_rows(.id = "ID") %>%
  select(-group)

# Calculate confidence intervals with confint().
CI <-  
  model_fit_list %>%
  future_map(confint, level = 0.95) %>%
  future_map(as_tibble, rownames = "term") %>%
  bind_rows(.id = "ID")

# Merge estimates and confidence intervals by formula number (ID) and term.
results_df <- formula_df %>%
  rownames_to_column(var = "ID") %>%
  inner_join(est, by = "ID") %>%
  inner_join(CI, by = c("ID", "term")) %>%
  select(-ID, -effect, -std.error, -statistic) %>%
  arrange(fgroup, formula, term)

# Stop timer.
toc()
```

Display the results.

```{r, fig.cap="Stepwise processing with merge."}
results_df %>% knitr::kable(digits = 4)
```

## Use a single pipeline

Since the results showed the two methods for producing confidence intervals are
mostly equivalent (except for the Intercept, for some reason), we will now use a 
single pipeline, without running `confint()`, so we can compare performance.

```{r}
# Start timer.
tic()

# Fit the models with lmer() and extract the estimates and confidence intervals 
# with broom.mixed::tidy() using future_map() for multicore processing.
results_df <- formula_df %>% 
  mutate(model = future_map(formula, lmer, data = data_df)) %>%
  mutate(est = future_map(model, tidy, conf.int = TRUE, conf.level = 0.95)) %>%
  select(-model) %>% unnest(cols = everything()) %>%
  filter(is.na(group)) %>% 
  select(-group, -effect, -std.error, -statistic) %>%
  arrange(fgroup, formula, term)
  
# Stop timer.
toc()
```

Display the results.

```{r, fig.cap="Single pipeline processing."}
results_df %>% knitr::kable(digits = 4)
```
