---
title: "Purrr and Broom Example"
author: "Brian High"
date: "11/1/2019"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Fit multiple models using purrr and broom

See: https://r4ds.had.co.nz/many-models.html

## Setup

```{r}
# Load packages.
pacman::p_load(datasets, tibble, dplyr, tidyr, purrr, broom, modelr, knitr)

# Get the dataset.
data(mtcars)
```

## Prepare data

```{r}
# Import the dataset into a nested tibble. Split car name into make and model.
df <- as_tibble(mtcars %>% 
                         mutate(model = row.names(mtcars))) %>% 
  mutate(make = gsub('^(\\w+) .*$', '\\1', model),
         model = gsub('^\\w+ (.*)$', '\\1', model)) %>% 
  group_by(make, model) %>% arrange(make, model)
```

### View data

* Use `kable()` from the `knitr` package.

```{r, render='asis'}
# Examine the dataset.
kable(df %>% head(10), format = 'markdown')
```

\pagebreak

### Nest data

Subset the data by whether or not the "make" is an American car maker. Nest the
groups for easier automation of modeling.

* Use `group_by()` from the `dplyr` package.
* Use `nest()` from the `tidyr` package.

```{r}
# Identify American car makers with: usa = TRUE.
df <- df %>% 
  mutate(usa = ifelse(make %in% c('Hornet', 'Valiant', 'Duster', 'Merc', 
                                  'Cadillac', 'Lincoln', 'Chrysler', 'Dodge', 
                                  'AMC', 'Camaro', 'Pontiac', 'Ford'), 
                      TRUE, FALSE))

# Nest by 'usa'.
df <- df %>% group_by(usa) %>% nest()
```

## Fit models

Fit using `lm()` with multiple formulas for each nested group. 

* Use `formulas()` and `fit_with()` from the `modelr` package.
* Use `map()` from the `purrr` package.

```{r}
# Define formulas.
lm_formulas <- formulas(~mpg, 
                              ~ cyl, 
                              ~ cyl + disp, 
                              ~ cyl + disp + hp, 
                              ~ cyl + disp + hp + wt)

# Fit models.
df <- df %>% 
  mutate(model = map(.x = data, .f = ~fit_with(lm, lm_formulas, data = .x)))
```

## Get model summaries and estimates

Extract the summary information from the models. We need to use `lapply()`
with `map()` because we are using many models and `fit_with()` created `list`
output like `lapply()` would have. So we use `lapply()` to get "inside" of the
`list` output of `fit_with()`.

* Use `glance()` and `tidy()` from the `broom` package.
* Use `map()` from the `purrr` package.
* Use `lapply()` from `base` to apply `glance()` and `tidy()` to many models.

```{r}
# Extract model summaries with glance().
df <- df %>% 
  mutate(resid = map(model, ~lapply(.x, glance)))

# Extract model estimates with tidy().
df <- df %>% 
  mutate(est = map(model, ~lapply(.x, tidy)))
```

\pagebreak

### View model summaries for "usa"

* Use `unnest()` from the `tidyr` package to unpack the `list` columns.
* Use `mutate()` from the `dplyr` package to include the formula names.

```{r}
df.resid <- df %>% select(usa, resid) %>% unnest(resid) %>% 
  mutate(formula = as.character(lm_formulas)) %>% unnest(resid)
```

* Use `kable()` from the `knitr` package.

```{r, render='asis'}
kable(df.resid[, c(1:7, 11, 13)], format = 'markdown', digits = 4)
```

\pagebreak

### View model estimates for "usa"

* Use `unnest()` from the `tidyr` package to unpack the `list` columns.
* Use `mutate()` from the `dplyr` package to include the formula names.

```{r}
df.est <- df %>% select(usa, est) %>% unnest(est) %>% 
  mutate(formula = as.character(lm_formulas)) %>% unnest(est)
```

* Use `kable()` from the `knitr` package.

```{r, render='asis'}
kable(df.est, format = 'markdown', digits = 4)
```
