---
title: "Table1"
author: "Rachel Shaffer"
date: "1/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to Table 1

In epidemiology and related fields, a "Table 1" is a table that presents descriptive statistics on the population under study. Often, we want to understand how baseline characteristics of a study population are related to certain stratifying variables (an exposure, an outcome, etc)

## Example Table 1 demo, closely adapted from: https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html 

### Table1 using formula interface

First, we need to import the data and create the stratifying variables that we are interested in. 

```{r import and prep data, echo=FALSE}

# loading data from boot package 
library(boot) 
melanoma2 <- melanoma
 
# Create basic variables that we're interested in
melanoma2$status <- 
  factor(melanoma2$status, 
         levels=c(2,1,3),
         labels=c("Alive", # Reference
                  "Melanoma death", 
                  "Non-melanoma death"))

```

Then, we can create a first attempt at Table1. Here, we will look at status (alive, melanoma death, non-melanoma death, and overall) by sex, age, ulcer, and thickness. 

We can create a table1 with a formula, where the key descriptive variables are separated by a "+" sign and the stratifying variable comes after the "|"

```{r first table 1, echo=FALSE}

# loading package 
library(table1)

# first attempt with formula 
table1(~ factor(sex) + age + factor(ulcer) + thickness | status, data=melanoma2)

```

The main problem with this table is that the descriptive labels don't look great. So we can manually add labels & units to them: 

```{r adding labels, echo=FALSE}

# adding labels & units 
melanoma2$sex <- 
  factor(melanoma2$sex, levels=c(1,0),
         labels=c("Male", 
                  "Female"))
 
melanoma2$ulcer <- 
  factor(melanoma2$ulcer, levels=c(0,1),
         labels=c("Absent", 
                  "Present"))

label(melanoma2$sex)       <- "Sex"
label(melanoma2$age)       <- "Age"
label(melanoma2$ulcer)     <- "Ulceration"
label(melanoma2$thickness) <- "Thickness"

units(melanoma2$age)       <- "years"
units(melanoma2$thickness) <- "mm"



# updated attempt with formula 
table1(~ sex + age + ulcer + thickness | status, data=melanoma2, overall="Total")


```

This looks good, but what if we want further customization? For example, we could put another level of header for "death", to group the melanoma & non-melanoma groups? And we could change what output is reported for each continuous variable? 

### Table 1 using list format 

If so, we will need to expand beyond the basic "formula" interface and set up the table using a list format. 

First, we set up the list of labels for variables and the groups that we want to display in our table:

```{r setting up list for table, echo=FALSE}

# setting up list of labels for variables and groups 
labels <- list(
    variables=list(sex="Sex",
                   age="Age (years)",
                   ulcer="Ulceration",
                   thickness="Thickness (mm)"),
    groups=list("", "", "Death"))

# Remove the word "death" from the labels, since it now appears above
levels(melanoma2$status) <- c("Alive", "Melanoma", "Non-melanoma")

```

Then, we set up the strata (columns) that we want to include in our table. 

```{r setting up strata, echo=FALSE}

strata <- c(list(Total=melanoma2), split(melanoma2, melanoma2$status))

```

Next, we can use custom rendering to specify what we want each type of variable (continuous vs. categorical) to display

```{r custom rendering, echo=FALSE}

my.render.cont <- function(x) {
    with(stats.apply.rounding(stats.default(x), digits=2), c("",
        "Mean (SD)"=sprintf("%s (&plusmn; %s)", MEAN, SD)))
}
my.render.cat <- function(x) {
    c("", sapply(stats.default(x), function(y) with(y,
        sprintf("%d (%0.0f %%)", FREQ, PCT))))
}

```

And then we can print our final table: 

```{r updated table, echo=FALSE}

table1(strata, labels, groupspan=c(1, 1, 2),
       render.continuous=my.render.cont, render.categorical=my.render.cat)

```

## More examples

More details on this table and other examples can be found here: https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html 

## Potential ideas for workshop explorations

1) Change appearance of table 
2) Display different statistics for different variables 
3) Practice similar code on a different dataset, with different stratifying variables 
4) Add a column of p-values 

## Summary

Overall, I think this function provides a lot of customization options for Table1. The main problem is that it only produces HTML, which is not useful if we are creating tables to put in publications! 
