---
title: 'Data Cleaning in Base-R: WA WQI'
author: "Brian High"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  ioslides_presentation:
    fig_caption: yes
    fig_retina: 1
    fig_width: 5
    fig_height: 3
    keep_md: yes
    smaller: yes
    template: ../../../templates/ioslides_template.html
  html_document:
    template: ../../../templates/html_template.html
editor_options: 
  chunk_output_type: console
---

```{r header, include=FALSE}
# Filename: get_wa_wqi_2013_with_base_R.Rmd
# Copyright (c) University of Washington
# License: MIT https://opensource.org/licenses/MIT (See LICENSE file.)
# Repository: https://github.com/deohs/coders
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

```{r get_template, include=FALSE}
# Copy template from package folder to work around Pandoc bug on Windows.
# If file is missing from template_dir, then first rendering attempt may fail.
template_pkg <- file.path(system.file(package = 'rmarkdown'), 
                          'rmd', 'ioslides', 'default.html')
if (.Platform$OS.type == "windows" & length(grep(' ', template_pkg) > 0)) { 
  template_dir <- file.path('..', '..', '..', 'templates')
  dir.create(template_dir, showWarnings = FALSE)
  template_loc <- file.path(template_dir, 'ioslides_template.html')
  if (!file.exists(template_loc)) {
    file.copy(template_pkg, template_loc, copy.mode = FALSE)
  }
}
```

## Using Base-R to Clean WA WQI 2013

Read the 2013 WA WQI file and parse coordinates into `lat` and `lon` using base-R.

Today's example demonstrates these objectives:

* Use a public dataset freely available on the web.
* Try various ways to clean up a dataset using only "base-R" functions.
* Use "regular expressions" to simplify data manipulation.
* Use "literate programming" to provide a reproducable report.
* Use a consitent coding [style](https://google.github.io/styleguide/Rguide.xml).
* Share code through a public [repository](https://github.com/deohs/coders) to facilitate collaboration.

The code and this presentation are free to share and modify according to the 
[MIT License](https://github.com/deohs/coders/blob/master/LICENSE).

## Read the data

All five variations of cleanup will use the same dataset imported with:

```{r}
url <- 'https://data.wa.gov/api/views/h7j9-vgr3/rows.csv?accessType=DOWNLOAD'
wa_wqi <- read.csv(url)
str(wa_wqi, vec.len = 1)
```

## Variation 1: strsplit and sapply

```{r}
v1_wa_wqi <- wa_wqi
coords <- as.data.frame(t(sapply(strsplit(trimws(
  gsub('[^0-9. -]', '', v1_wa_wqi$Location.1)), ' '), as.numeric)))
names(coords) <- c('lon', 'lat')
v1_wa_wqi <- cbind(v1_wa_wqi, coords)

head(v1_wa_wqi[, c('STATION', 'OVERALLWQI.2013', 'lon', 'lat')])
str(v1_wa_wqi[ , c('lon', 'lat')])
```

## Variation 2: strsplit and lapply

```{r}
v2_wa_wqi <- wa_wqi
coords <- as.data.frame(do.call('rbind', strsplit(
    gsub('POINT |[()]', '', v2_wa_wqi$Location.1), ' ')), stringsAsFactors = FALSE)
coords <- as.data.frame(lapply(coords, as.numeric), col.names = c('lon', 'lat'))
v2_wa_wqi <- cbind(v2_wa_wqi, coords)

head(v2_wa_wqi[, c('STATION', 'OVERALLWQI.2013', 'lon', 'lat')])
str(v2_wa_wqi[ , c('lon', 'lat')])
```


## Variation 3: gsub (twice)

```{r}
# We can just copy, paste, and modify our gsub() code, but what a nasty habit!
v3_wa_wqi <- wa_wqi
regex <- c(lon = '^.*\\(([0-9.-]+) .*$', lat = '^.*\\(.* ([0-9.-]+)\\)$')
v3_wa_wqi$lon <- as.numeric(gsub(regex[['lon']], '\\1', v3_wa_wqi$Location.1))
v3_wa_wqi$lat <- as.numeric(gsub(regex[['lat']], '\\1', v3_wa_wqi$Location.1))

head(v3_wa_wqi[, c('STATION', 'OVERALLWQI.2013', 'lon', 'lat')])
str(v3_wa_wqi[ , c('lon', 'lat')])
```

## Variation 4: gsub and lapply

```{r}
# We avoid the dreaded copy-and-paste with lapply(). But is it worth it?
v4_wa_wqi <- wa_wqi
regex <- c(lon = '^.*\\(([0-9.-]+) .*$', lat = '^.*\\(.* ([0-9.-]+)\\)$')
v4_wa_wqi[, names(regex)] <- lapply(names(regex), function(x) {
  as.numeric(gsub(regex[[x]], '\\1', v4_wa_wqi$Location.1)) })

head(v4_wa_wqi[, c('STATION', 'OVERALLWQI.2013', 'lon', 'lat')])
str(v4_wa_wqi[ , c('lon', 'lat')])
```

## Variation 5: cbind and read.table

```{r}
# This one accomplishes the split with less code -- in just one statement.
v5_wa_wqi <- wa_wqi
v5_wa_wqi <- cbind(v5_wa_wqi, read.table(
  text = gsub('^POINT \\((.*)\\)$', '\\1', v5_wa_wqi$Location.1), 
  sep = ' ', col.names = c('lon', 'lat')))

head(v5_wa_wqi[, c('STATION', 'OVERALLWQI.2013', 'lon', 'lat')])
str(v5_wa_wqi[ , c('lon', 'lat')])
```

## Compare results

Are the all of the variables indentical in all of the variations? Perform a 
pairwise comparison of the 1st variation against each of the others to find out.

```{r}
c(identical(v1_wa_wqi, v2_wa_wqi),
  identical(v1_wa_wqi, v3_wa_wqi),
  identical(v1_wa_wqi, v4_wa_wqi),
  identical(v1_wa_wqi, v5_wa_wqi))
```

Since we generally do not like to copy and paste code over and over, we may 
prefer this alternative:

```{r}
sapply(list(v2_wa_wqi, v3_wa_wqi, v4_wa_wqi, v5_wa_wqi), 
       function(x) identical(v1_wa_wqi, x))
```

## Excercises

1. In Var. 1, what data structure does `strplit()` return? What did `sapply()` 
   do and why was `t()` needed?
2. In Var. 2, what does `as.data.frame(do.call('rbind', ...` do and what is 
   the tidyverse replacement for this idiom? What does `as.data.frame(lapply( ...`
   do and what tidyverse function will accomplish this?
3. In Var. 3, what are the two regular expressions doing. Explain each piece of
   each expression and what it matches.
4. In Var. 4, the method of Var. 3 was implemented using `lapply()`. Aside from
   wanting to avoid copy-and-paste, why would anyone want to go to the trouble?
5. In Var. 5, we did not need to use `as.numeric()` to set the correct variable 
   types as we did on the other 4 variations. How was this possible?
6. Are these results identical with those produced with tidyverse functions 
   `mutate()` and `separate()`?

```{r, eval=FALSE}
library(tidyverse)
v6_wa_wqi <- read_csv(url) %>% 
  mutate(Location.1 = gsub('POINT |[()]', '', `Location 1`)) %>%
  separate(col = Location.1, into = c('lon', 'lat'), sep = ' ', convert = TRUE)
identical(as_tibble(v1_wa_wqi[, c('lat', 'lon')]), v6_wa_wqi[, c('lat', 'lon')])
```
